Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-04 surfaces-write-up/surface-write-up.tex
I don't recall making the change it says I made... oops...
deal with peter's merge conflict...
responding to todos

\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 1}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\todo{transition}

@@

\todo{transition}

@@

\todo{Include summary of what each of the sections do.}
\todo{Incorporate the following section into the $\bp^{m}$ section}

@@

\begin{example}
\label{eg:radical}
Consider $\frac{-1}{5}V(x_0^2 + x_1^2 + x_2^2) + \frac{1}{7}V(x_0^2 + x_1^2 +
x_3^2) + \frac{1}{17}V(x_0^2 + x_2^2 + x_3^2) - \frac{1}{596}V(x_1^2 + x_2^2 +
x_3^2)$. In degree $355216 = 5 \cdot 7 \cdot 17 \cdot 596$, this has dimension
$10$. 
\todo{Clarify what this example means by the lattice approach}
However, we possibly use the lattice approach, as all lower degrees only have
dimension 1. In fact, we may construct similar examples using any set of degree
$n$ divisor with very small degree.
\end{example}
\todo{construct class of additional examples}

@@

We claim such a $\beta_i$,
\todo{Aaron: Shouldn't we say something above like, the kernel
	is bounded by the maximum of something over all $i$,
	so for the remainder of the proof, we fix an $i$.
	Also, I think it would help if things
	that depend on $i$, such as $E,\sigma,\lambda$
have an $i$ subscript.

Peter: Better (on the first comment)? As for the second, that might be a little tricky, because we already have $E_\mu$, so we would have to do something like $E_{i,\mu}$ (which we could of course do...)

Aaron: Does $E_\mu$ depend on $i$? I thought it didn't, but if it does, then we should definitely do $E_{i,\mu}$ since I didn't
realize it depends on $i$. If $E_\mu$ doesn't depend on $i$, then we can just write $E_i$, and $E_\mu$,
and there is no danger of confusion because the subscripts live in different categories.
\\
Peter:  Yes, essentially everything in the remainder of this proof depends on $i$, e.g.: $\omega$, $\lambda$ (depends on $i$ as well as the choice of $r_{i,j}$'s), $E$, all $E_\mu$'s, $F$, $F_\mu$'s; I think that is everything.  I do not believe it would be helpful to add subscripts to all of these.  Instead maybe it would be better just to have one sentence that very clearly says everything in the following proof fixes $i$ and depends on $i$.

Aaron: How does $E_\mu$ depend on $i$? If you'd prefer to just say everything depends on $i$, I'm ok with it, though I still think including $i$ subscripts would be slightly
preferable. Perhaps just say near the beginning a list of all the things that will depend on $i$. I don't think you can say "everything for the remainder of the proof depends on i", because that's not true, e.g., $\ell_j$ for $j \neq i$ doesn't depend on $i$, though of course this is very confusing because you write $\ell_i$ in the definition of $\deg E$. (Won't this be more difficult than just writing a supscript or superscript $i$ though?)
\\
Peter: Where does $\ell_j$ appear?  It may be a typo... I can't remember...  I think there are only $\ell_i$'s and $\ell_{i,j}$'s.  Anyway, what I meant was every new variable defined after this point depends on $i$, not every variable used.

Aaron: You're right, I was confused. It is just $l_{i,j}$.
}
can be expressed as a polynomial $\beta_i(f_1,f_2,f_3,f_4)$ with 
$\beta_i(f_1,f_2,f_3,f_4) = f_i$
as elements of $\bk[\Sigma],$ which is possible as $f_1,f_2,f_3,f_4$ generate $\Sigma,$ and hence generate $\bk[\Sigma]$. Furthermore, these
generate all relations, since they allow us to reduce any monomial $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$
\todo{Aaron: Should $r_i$ be $c_i$ or should $c_i$ be $r_i$?}
\todo{Aaron: Am I misremembering, or did we agree on the phone that you would put $i$ subscripts on all the things that depend on $i$?
If not, weren't you at least going to say somewhere which things depend on $i$?}
 whenever $i > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

We now wish to bound the degree of generation the ideal of relations generated by these.  To do this, we first fix $i$ and a relation of the form $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0.$ Note that there are no relations for $i\in \{1, 2, 3, 4\}$ as $\beta_i = f_i$. Our goal is to find a relation that the divides this, but whose degree we can bound. We do so by considering the lattice points corresponding to the monomials appearing in the relation $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$, and finding some $\lambda_\omega\in \Sigma$
\todo{How can you define $\lambda_\omega$ without yet defining $\omega$?
Also, you only define $\lambda$ later, not $\lambda_\omega$. I don't even think you define $\omega$ later.
} that we can factor out of each monomial, to get a relation below a bounded degree.  The following argument makes this precise:

@@

If $i\in \Te$ define $r_i := \frac{1}{\kappa_i}$.  If $i\in \Tp$ choose $r_{i,j}\in \bq_{\ge 0}$ for all $j\in \Tm$ such that $\sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1$; similarly, if $i\in \Tm$ choose $r_{j,i}\in \bq_{\ge 0}$ for all $j\in \Tp$ such that $\sum_{j\in \Tm} r_{j,i}\kappa_{j,i}'' = 1$.  For $j\ne i$, define $r_j := 0$ and for all pairs $(j,k)$ so that $j \neq i$ and $k \neq i$ define $r_{j,k} := 0$.
Define $E$ by
\begin{equation}\label{eqn:hirz-E-translation}
	E := \sum_{j\in \Te} (s_j - \lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} (s_{j,k} - \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}.
\end{equation}
Define
\begin{equation}\label{eqn:defn-lambda}
	\lambda := \sigma - E = \sum_{j\in \Te} (\lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} ( \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}\in \Sigma.
 \end{equation}
Note that
\[
	\deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
\]
where $\rho$ is as in Equation \ref{eqn:def-sigma}.
\todo{Aaron: Is one of these terms corresponding to $E$ and the other to $\lambda$? If so, can you say which?}

Let $M_i$ be the set of monomials terms of $\beta_i = \beta_i(f_1, \ldots f_4)$.  Let $\mu = \prod_{j=1}^4
{f_j}^{h_j}\in M_i$ and consider the lattice point 
\[
	\sigma_\mu = (d, c_1 + h_1, \ldots, c_4+h_4, c_5, \ldots, c_n).
\]
where $d = \deg \sigma$.
Define 
\[
	E_\mu = \sigma_\mu - \lambda.
\]
A quick computation deduces that $E_\mu$ lies in $\Sigma$ (using the condition on the $r_{i,j}$'s and $r_{j,i}$'s \todo{should one of these be $r_i$'s?}) and has the same degree as $E$.

Let $F\in R_D$
\todo{Aaron: Didn't we agree you would write these as $\psi(E)$ or something like that?} be the monomial given by the lattice point $E$ and $F_\mu \in R_D$ be the monomial given by the lattice point $E_\mu$ for each $\mu \in M_i$.  
Then $F - \sum_{\mu \in M_i} F_\mu = 0$ is a relation %in degree
%\[
%	\deg(F) = \deg(E)
%\]
which divides $(z_i - \beta_i) \prod_{j = 1}^n z_j^{c_j}$. 
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
	\]

>>>>

In section 2, we discuss background knowledge and present some useful lemmas.  In section 3 we focus on projective spaces: $\mathbb{P}^n$ and prove the following two results:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 1}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

@@

\noindent
In section 4, we shift our attention to Hirzebruch Surfaces and prove the following theorem:

@@

\noindent
Together, Theorems \ref{thm:proj-generators-relations} and \ref{thm:hirz-generators-relations} provide bounds on generators and relations for the canonical ring of $\mathbb{Q}$-divisors of all minimal rational surfaces.  Section 5 concludes the paper and discusses further questions.

\todo{Incorporate the following section into the $\bp^{m}$ section
\\
Peter: I'm not sure what this refers to...  Can you be more specific}

@@

\begin{example}
\label{eg:radical}
Consider $\frac{-1}{5}V(x_0^2 + x_1^2 + x_2^2) + \frac{1}{7}V(x_0^2 + x_1^2 +
x_3^2) + \frac{1}{17}V(x_0^2 + x_2^2 + x_3^2) - \frac{1}{596}V(x_1^2 + x_2^2 +
x_3^2)$. In degree $355216 = 5 \cdot 7 \cdot 17 \cdot 596$, this has dimension
$10$. 
\todo{Clarify what this example means by the lattice approach 
\\
Peter: Maybe we should put this after the ensuing proofs.  That way, rather than saying the Lattice Approach, we could say that this example illustrates how the prveious method requires us to add ghost points (and is false otherwise).}
However, we possibly use the lattice approach, as all lower degrees only have
dimension 1. In fact, we may construct similar examples using any set of degree
$n$ divisor with very small degree.
\end{example}
\todo{construct class of additional examples}

@@

We claim such a $\beta_i$,
can be expressed as a polynomial $\beta_i(f_1,f_2,f_3,f_4)$ with 
$\beta_i(f_1,f_2,f_3,f_4) = f_i$
as elements of $\bk[\Sigma],$ which is possible as $f_1,f_2,f_3,f_4$ generate $\Sigma,$ and hence generate $\bk[\Sigma]$. Furthermore, these
generate all relations, since they allow us to reduce any monomial $\prod_{j =
1}^n f_j^{r_j}$ to the canonical form where $r_j = - \lfloor d \alpha_i
\rfloor$
\todo{Aaron: Am I misremembering, or did we agree on the phone that you would put $i$ subscripts on all the things that depend on $i$?
If not, weren't you at least going to say somewhere which things depend on $i$?
\\
Peter: I still don't entirely agree that explaining the dependency is important.  What I mean by this is that the proof fixes $i, \sigma,$ etc. and proves the desired result.  Proving it for each possible choice of course proves the general result; however, it doesn't seem important to me whether or not the intermediate variables are the same in the proof for $i'$ and $\sigma'$ than for $i$ and $\sigma$.  Either way the proof is valid.  This is why I have been somewhat resistant (sorry if annoyingly so...)to adding them in.

On a side note, in some cases actually writing them in would be cumbersome, e.g. replacing $s_{j,k}$ with $s_{j,k,\sigma}$ (since there is one choice of $s_{j,k}$ for every $\sigma$.

If you want, can you tell me exactly which variables would be helpful in understanding the proof to know their dependencies?
}
 whenever $j > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

We now wish to bound the degree of generation the ideal of relations generated by these.  To do this, we first fix $i$ and a relation of the form $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0.$ Note that there are no relations for $i\in \{1, 2, 3, 4\}$ as $\beta_i = f_i$; thus we restrict attention to $i \ge 5$. Our goal is to find a relation that the divides this, but whose degree we can bound. We do so by considering the lattice points corresponding to the monomials appearing in the relation $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$, and finding some $\lambda\in \Sigma$
that we can factor out of each monomial, to get a relation below a bounded degree.  The following argument makes this precise:

@@

If $i\in \Te$ define $r_i := \frac{1}{\kappa_i}$.  If $i\in \Tp$ choose $r_{i,j}\in \bq_{\ge 0}$ for all $j\in \Tm$ such that $\sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1$; similarly, if $i\in \Tm$ choose $r_{j,i}\in \bq_{\ge 0}$ for all $j\in \Tp$ such that $\sum_{j\in \Tm} r_{j,i}\kappa_{j,i}'' = 1$.  For $j\ne i$, define $r_j := 0$ and for all pairs $(j,k)$ so that $j \neq i$ and $k \neq i$ define $r_{j,k} := 0$.
Define $E$ by
\begin{equation}\label{eqn:hirz-E-translation}
	E := \sum_{j\in \Te} (s_j - \lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} (s_{j,k} - \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}.
\end{equation}
Define
\begin{equation}\label{eqn:defn-lambda}
	\lambda := \sigma - E = \sum_{j\in \Te} (\lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} ( \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}\in \Sigma.
 \end{equation}
Note that
\[
	\deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
\]
where $\rho$ is as in Equation \ref{eqn:def-sigma}.
\todo{Aaron: Is one of these terms corresponding to $E$ and the other to $\lambda$? If so, can you say which?\\
Peter: No, one corresponds to $r_j$ and one to the fractional part of $s_j - r_j$.}

Let $M_i$ be the set of monomials terms of $\beta_i = \beta_i(f_1, \ldots f_4)$.  Let $\mu = \prod_{j=1}^4
{f_j}^{h_j}\in M_i$ and consider the lattice point 
\[
	\sigma_\mu = (d, c_1 + h_1, \ldots, c_4+h_4, c_5, \ldots, c_n).
\]
where $d = \deg \sigma$.
Define 
\[
	E_\mu = \sigma_\mu - \lambda.
\]
A quick computation deduces that $E_\mu$ lies in $\Sigma$ 
%(using the condition on the $r_{i,j}$'s and $r_{j,i}s$) \todo{should one of these be $r_i$'s?}.  \\ Peter: No, but I'm not sure if the parenthetical is helpful anyway.  $r_i$ isn't define or used (hopefully...) anywhere.) 
and has the same degree as $E$.

Let $F = \Psi(E) \in R_D$
\todo{Aaron: Didn't we agree you would write these as $\psi(E)$ or something like that? 
\\
Peter: Okay, I'm still a little unhappy with with the identification of lattice points in $\Sigma$ with monomials (since $\Psi$ maps monomials in a polynomial ring to monomials in $R_D$), but I made the change} and $F_\mu = \psi(E_\mu) \in R_D$ for each $\mu \in M_i$.  
Then $F - \sum_{\mu \in M_i} F_\mu = 0$ is a relation %in degree
%\[
%	\deg(F) = \deg(E)
%\]
which divides $(z_i - \beta_i) \prod_{j = 1}^n z_j^{c_j}$. 
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
	\]
